Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-15 paper_io_sim.tex
rewording related works

        %To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
        The goal of the page cache is to ensure that most I/O operations occur entirely
        within memory space rather than synchronously to and from storage devices.
        All disk-backed data located in the page cache are partitioned into many small pages.
        When a read operation is initiated, the kernel checks whether the required pages are in memory.
        If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
        the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss}).
        Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
        writethrough and writeback.
        If direct~I/O is enabled for a given file system, all I/O operations bypass the
        page cache for writes, writing data directly to disk. In contrast, writethrough enables the page cache
        for file system writes, but the data is written synchronously from memory to disk. Writeback extends writethrough
        by enabling asynchronous writes to disk, meaning that a write call will return once
        the data has been written to memory.
        The writeback strategy is considered to outperform writethrough as well as
        direct I/O as it delays disk writes to perform a bulk write at a later time
        \cite{linuxdev3rd2010}.

        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes expired dirty pages, which remain dirty in
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been developed and proposed
        \cite{owda2014comparison}.
        The Linux kernel uses a two-list strategy to flag pages for eviction.
        The two-list strategy is based on a least recently used (LRU) policy
        and uses an active and inactive list in its implementation.
        If accessed pages are not in the page cache, they are added to the inactive list.
        Should pages located on the inactive list be accessed, they will be moved from
        the inactive to the active list.
        The lists are also kept balanced by moving pages from the active list
        to the inactive list when the active list grows too large.
        Thus, the active list only contains pages which are accessed more than once
        and not evictable, while the inactive list includes pages accessed once only,
        or pages that have be accessed more than once but moved from the active list.
        Both lists operate using LRU eviction policies meaning the data that has
        not be accessed recently will be moved first.

>>>>

        %To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}
        Page cache offsets the cost of disk I/O by enabling I/O to occur directly from memory.
        When a file is first loaded into memory, the file is read from disk and loaded into the page cache as
        a series of pages. Subsequent reads to any of the files pages located in memory will result in
        a \textit{cache hit}, meaning the I/O can occur directly from without disk involvement.
        Any accessed page not loaded in memory results in a \textit{cache miss}, resulting in
        the page being read directly from disk.

        Written pages can also contribute to future application cache hits. When use of page cache
        is enabled for a given filesystem, through the enabling of writeback or writethrough cache,
        all written pages are written first to page cache, prior to being written to disk.
        Accessing of these written pages may result in cache hits, should the pages remain in memory.

        The kernel may also provide improved write performance if writeback cache is enabled. Unlike writethrough,
        where the data is synchronously written from memory to disk, writeback enables asynchronous writes.
        In other words, with writeback, the application may proceed with execution once the data has been
        written to memory, even if it has not yet been materialized to disk.  
        The writeback strategy is considered to outperform writethrough as well as
        direct I/O (page cache bypassed for I/O) as it delays disk writes to perform a bulk write at a later time
        \cite{linuxdev3rd2010}.

        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes expired dirty pages, which remain dirty in
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been developed and proposed
        \cite{owda2014comparison}.

        The Linux kernel uses a two-list strategy to flag pages for eviction.
        The two-list strategy is based on a least recently used (LRU) policy
        and uses an active and inactive list in its implementation.
        If accessed pages are not in the page cache, they are added to the inactive list.
        Should pages located on the inactive list be accessed, they will be moved from
        the inactive to the active list.
        The lists are also kept balanced by moving pages from the active list
        to the inactive list when the active list grows too large.
        Thus, the active list only contains pages which are accessed more than once
        and not evictable, while the inactive list includes pages accessed once only,
        or pages that have be accessed more than once but moved from the active list.
        Both lists operate using LRU eviction policies meaning the data that has
        not be accessed recently will be moved first.
